---
title: "Useful R Stuff"
output:
  html_document:
    css: "style.css"
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false

---
<script src="fold_rmd.js"></script>


<!-- <div class = "fold s o"></div> to hide "s"ource or "o"utput or both -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r echo = FALSE, results = "hide", message = FALSE}
library(tidyverse)

```


# Working with dataframes

## thinkr::peep() - see intermediate stages of piped dfs

```{r}

iris %>% 
  thinkr::peep(head) %>% 
  rename(species = Species)

```
```{r}
iris %>% 
  thinkr::peep(head(., n = 7)) %>% 
  rename(species = Species)
```


## janitor - clean dirty data, tabulating tools, adornments

Clean up column names
```{r results = "hide", message = FALSE}
library(janitor)
```

```{r}
# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")

test_df %>%
  clean_names()

test_df %>% 
  clean_names(case = "small_camel")
```

Compare two or more DFs to see if they have the same types of columns and can be rbound
```{r}
df1 <- data.frame(a = 1:2, b = c("big", "small"))
df2 <- data.frame(a = 10:12, b = c("medium", "small", "big"), c = 0, stringsAsFactors = TRUE) # here, column b is a factor
df3 <- df1 %>%
  dplyr::mutate(b = as.character(b))

compare_df_cols(df1, df2, df3)
```

## janitor::tabyl - compare combinations of varibales like xtab

```{r}
mtcars %>%
  tabyl(gear, cyl)
```

You can "adorn" tabuls with totals, percentages, and more

```{r}
mtcars %>%
  tabyl(gear, cyl) %>%
  adorn_totals("col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns() %>%
  adorn_title()
```

[More on tabyls](http://sfirke.github.io/janitor/articles/tabyls.html)



## diffobj - compare two dfs (or other objects) to see what's changed between them

NB: in RStudio this will open a colorful thing in the Viewer that's easy to read

<div class = "fold o">
```{r}
library(diffobj)
a <- b <- matrix(1:100, ncol=2)
a <- a[-20,]
b <- b[-45,]
b[c(18, 44)] <- 999
diffPrint(target=a, current=b)
```
</div>


<div class = "fold o">
```{r}
diffChr(letters[1:3], c("a", "B", "c"))
```
</div>



## fuzzyjoin: Join dfs on inexact matching

Among other options:

- Numeric values that are within some tolerance (difference_inner_join)
- Strings that are similar in Levenshtein/cosine/Jaccard distance, or other metrics from the stringdist package (stringdist_inner_join)
- A regular expression in one column matching to another (regex_inner_join)

# Summarizing
## skimr::skim() - nice summary statistics for dfs and the like

<div class = "fold o">
```{r}
library(skimr)
iris %>% 
  group_by(Species) %>% 
  skim() %>% 
  print() # print() only necessary so we don't get a kable table in this document
```
</div>

---

```{r}
iris %>% 
  skim() %>% 
  summary() %>% 
  print() # print() only necessary so we don't get a kable table in this document
```

---

<div class = "fold o">
```{r}
iris %>% 
  skim(Sepal.Length, Petal.Length) %>%  # tidyselect columns to summarize
  print() # print() only necessary so we don't get a kable table in this document
```
</div>

---

<div class = "fold o">
You can do cool stuff in the pipeline with it

```{r}
iris %>%
  skim() %>%
  dplyr::filter(numeric.sd > 1) %>% # "numeric" is the variable type that we're targeting
  print()
```

</div>

# Plotting
## Khroma - color scales that are nice and colorblind friendly

https://packages.tesselle.org/khroma/


```{r}
library(khroma)
bright <- colour("bright")
bright(7)
```

<div class="fold o">
```{r}
# Preview the theme
plot_scheme(bright(7), colours = TRUE)
```
</div>

Use with ggplot2

```{r}

data(mpg, package = "ggplot2")

ggplot2::ggplot(data = mpg) +
  ggplot2::aes(x = displ, y = hwy, colour = class) +
  ggplot2::geom_point() +
  scale_colour_bright()
```





## ggiraph - interactive graphs even in Rmd

This can do a whole lot of cool stuff more than this simple example.

More info: <https://davidgohel.github.io/ggiraph/>


```{r}
library(ggplot2)
library(ggiraph)
data <- mtcars
data$carname <- row.names(data)

gg_point = ggplot(data = data) +
    geom_point_interactive(aes(x = wt, y = qsec, color = disp,
    tooltip = carname, data_id = carname)) + 
  theme_minimal()

girafe(ggobj = gg_point)
```



## Patchwork - composing multiple plots together

Setting up some plots we can stitch together
<div class = "fold s">
```{r messages = FALSE}
library(patchwork)
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```
</div>


<div class = "fold o">

```{r}
p1 + p2 # combine two plots
```
</div>


<div class = "fold o">
```{r}
p1 + p2 + p3 + p4 # by default it makes a square grid, filling rowwise
# plot_layout()
p1 + p2 + p3 + p4 + plot_layout(nrow = 3, byrow = FALSE)
```
</div>


Stacking and packing plots:
<div class = "fold o">
```{r}
p1 / p2 # / stacks one on top of the other
p1 | (p2 / p3) # | plots side by side.  Note the parens
```
</div>

<div class = "fold o">
```{r}
p1 + p2 +
  plot_annotation(
    title = 'This is a title',
    caption = 'made with patchwork',
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5))
  )
```
</div>

You can define your own layout using letters, with relative sizing
<div class = "fold o">
```{r}
layout <- "
##BBBB
AACCDD
##CCDD
"
p1 + p2 + p3 + p4 + 
  plot_layout(design = layout)
```

</div>




It gets a lot more [complicated and cool](https://patchwork.data-imaginist.com/).  See [here](https://patchwork.data-imaginist.com/articles/guides/multipage.html) for how this package can also avoid plots "jumping" around as you switch slides.

## Paginate faceted graphs and save as a single PDF

You can manually select which page to print:

```{r}
USArrests %>% # dataset on US arrests
  rownames_to_column("State") %>%  # This is just so we can see the state names
  ggplot(aes(x = UrbanPop, y = Murder)) +
  geom_point() +
  ggforce::facet_wrap_paginate(~State, nrow =3, ncol = 3, page = 2)
```
Or you can be fancy and do a loop.  If you know the number of pages ahead of time (which you can get with `ggforce::n_pages(p)`), you can just loop through those:

```{r eval = FALSE}
cairo_pdf('filename.pdf', width = 11, height = 8.5, onefile = TRUE) #start building pdf
for (i in 1:9) {
  print( # don't forget this
    USArrests %>% # dataset on US arrests
    rownames_to_column("State") %>%  # This is just so we can see the state names
    ggplot(aes(x = UrbanPop, y = Murder)) +
    geom_point() +
    ggforce::facet_wrap_paginate(~State, nrow =3, ncol = 3, page = 2)
  )
}
dev.off() # end building pdf
```

Or you can calcuate the number of pages on the fly: (https://stackoverflow.com/questions/48544039/how-to-save-output-from-ggforcefacet-grid-paginate-in-only-one-pdf)



# Package and project management


## Package development workflow
<https://r-pkgs.org>: the definitive book by Wickham and Bryan, which explains lots of the stuff below

```{r eval = FALSE}

devtools::load_all() # simulates installing and loading your package
devtools::document() # updates all the documentation and stuff via roxygen2
devtools::check() # run some checks when you think you're ready to release

usethis::use_version(which = c("major", "minor", "patch", "dev")) # update version number.  `which` is what level of x.y.z.dev to increment

```


## usethis


```{r eval = FALSE}
# Creating
create_package()
create_project()

# Checking
proj_sitrep() # show wd, proj dir, etc

# Development
use_package() # import packages into DESCRIPTION

# Include important things
use_pipe()
use_tibble()

# Git
use_git()
use_github()
git_sitrep()

```
Also has git/github functions and other config stuff



# Modeling

## TidyModels
<https://www.tidymodels.org>; has a bunch of subpackages, including broom



# Data exploration
## DataExplorer
<https://boxuancui.github.io/DataExplorer/>

<div class = "fold o">
```{r message=FALSE}
library(DataExplorer)
plot_bar(ggplot2::diamonds) # Plot frequency of levels in each factor
```
```{r}
plot_histogram(ggplot2::diamonds) # Histogram for all continuous variables
```
```{r}
plot_boxplot(ggplot2::diamonds, by = "cut")
```
</div>

# Regex

## RegExplain add-in

Select a variable name in your script, then Addins > RegExplain.  You may want to type out dataframe$variable so it selects just that column.

Also Addins > RegExplain Cheatsheet has lots of helpful stuff


## rex - "friendly regular expressions"

A bunch of shortcuts to make regexprs more human-readable

```{r message = FALSE}
library(rex)
```

"shortcuts": Commonly used character classes and regular expressions. These shortcuts are substituted inside rex calls.

NB: to have these shortcuts autocomplete in Rstudio, you must first run `rex::rex_mode()`

<div class="fold o">
```{r}
rex::shortcuts
```
</div>


Capture groups:

```{r}
x <- c("a=1", "b=3", "a=9", "c=2", "b=4", "a=2")

matches <- re_matches(x,
                      rex(
                        capture(name="name", letter),
                        "=",
                        capture(name="value", digit)
                      ))
matches
```

More capture groups:
<div class = "fold s o">
```{r}
# extract 2,9,22,22. These characters are between the / and after the first character within the /.

mystrings <- c("X2/D2/F4",
               "X10/D9/F4",
               "X3/D22/F4",
               "X9/D22/F9")

matches <- re_matches(mystrings,
                      rex(
                        "/",
                        any,
                        capture(name = "numbers", digits)
                      )
)
as.numeric(matches$numbers)

```
</div>

Even more more capture groups
<div class = "fold s o">
```{r}
j <- "What kind of cheese isn't your cheese? (wonder) Nacho cheese! (groan) (Laugh)"

matches <- rematch2::re_match_all(j,
                      rex(
                        "(",
                        capture(name = "text", except_any_of(")")),
                        ")")) %>% unnest(text)
matches
```
</div>

`or()`
```{r}
re <- rex(or("A", "B", "C"))
str_match(re, "A")
str_match(re, "B")
str_match(re, "C")
```


Create character classes
```{r}
re <- rex("gr", one_of("a", "e"), "y") # outputs gr[ae]y
rematch2::re_match_all(c("gray", "grey"), re) %>% unnest()

re <- rex(none_of("a", "e" ,"i", "o", "u")) # [^aeiou]
rematch2::re_match_all(c("ay", "be", "cee"), re) %>% unnest()
```
Escape stuff:
```{r}
re <- rex(escape("[hello]")) # \[hello\]
str_match("[hello]yello", re)

# If using for a string that will define a character class, use character_class_escape() instead
```

Wildcards

```{r}
re <- rex(zero_or_more(digit))
str_extract_all("123abc456", re)

re <- rex(one_or_more(digit))
str_extract_all("123abc456", re)

re <- rex(maybe(digit))
str_extract_all("123abc456", re)
```



## rematch2 - like str_match(_all), except put it in a tidy tibble instead of a list

(May want to follow with `%>% unnest()` to expand the list-columns)


```{r}
name_rex <- paste0(
  "(?<first>[[:upper:]][[:lower:]]+) ",
  "(?<last>[[:upper:]][[:lower:]]+)"
)
notables <- c(
  "  Ben Franklin and Jefferson Davis",
  "\tMillard Fillmore"
)
not <- rematch2::re_match_all(notables, name_rex)

not

not %>% unnest(c(first, last))
```

# Sorting

## naturalsort - sort numbers without having to left-zero-pad them

```{r}
sprintf("hoge%s", 1:12) %>% sample(.) %>%
  naturalsort::naturalsort(.) %>% print()
```



# Etc

## Progress bars with `progress`

```{r eval = FALSE}
list_to_traverse <- 1:10

# Set up the progress bar
pb <- progress::progress_bar$new(total = 10,
                                 format = "  [:bar] :spin :percent (:current/:total) in :elapsedfull (eta: :eta) ",
                                 clear = FALSE)
purrr::walk(list_to_traverse,
            function(x) {
              Sys.sleep(.5)
              pb$tick() #call pb$tick() after each iteration
            })

beepr::beep(sound = "fanfare") # play a sound after
```

You can also do this inline in a one-sided formula (~).  NB that the entire one-sided formula after the ~ must be in {curly braces} and use semicolon to separate "lines"
```{r eval = FALSE}
purrr::walk(list_to_traverse,
            ~ {Sys.sleep(.5); pb$tick()}
            )
```

Package `asbutils` has a version of `progress_bar` with sane defaults:
```{r eval = FALSE}
pb <- asbutils::asb_progress(total = 10)
```




## datapasta - copy/paste data

```{r message = FALSE}
library(datapasta)
```

Paste as a tribble or as a df

| A | B | C | D | E |
|---|---|---|---|---|
| 1 | 4 | 5 | 7 | 9 |
| 8 | 5 | 8 | 0 | 8 |
| 5 | 6 | 8 | 9 | 0 |


```{r eval = FALSE}
datapasta::tribble_paste()
# Pastes the following
# tibble::tribble(
#      ~A, ~B, ~C, ~D, ~E,
#      1L, 4L, 5L, 7L, 9L,
#      8L, 5L, 8L, 0L, 8L,
#      5L, 6L, 8L, 9L, 0L
#      )

datapasta::df_paste()
# data.frame(
#            A = c(1L, 8L, 5L),
#            B = c(4L, 5L, 6L),
#            C = c(5L, 8L, 8L),
#            D = c(7L, 0L, 9L),
#            E = c(9L, 8L, 0L)
#    )
```

| A | B | C | D | E |
|---|---|---|---|---|

```{r eval = FALSE}
datapasta::vector_paste()
# Pastes the following
# c("A", "B", "C", "D", "E")

datapasta::vector_paste_vertical()
# c("A",
#      "B",
#      "C",
#      "D",
#      "E")
```

Get data from an existing table into usable format - useful for making a reprex/mwe!
```{r eval = FALSE}
paste_me <- tibble::tribble(
     ~A, ~B, ~C, ~D, ~E,
     1L, 4L, 5L, 7L, 9L,
     8L, 5L, 8L, 0L, 8L,
     5L, 6L, 8L, 9L, 0L
     )

dpasta(paste_me)
# Pastes the following
# tibble::tribble(
#      ~A, ~B, ~C, ~D, ~E,
#      1L, 4L, 5L, 7L, 9L,
#      8L, 5L, 8L, 0L, 8L,
#      5L, 6L, 8L, 9L, 0L
#      )

# dmdclip(paste_me) does the same as dpasta, but pastes to keyboard
```

Two useful Addins:
`Toggle Vector Quotes` highlight a c() expression and it either wraps or unwraps all elements with double quotes

`Fiddle Selection` does a bunch:

- If you highlight raw data (separated by commas, spaces, newlines), it will turn it into a c()
- If you select a c() expression, will toggle between horizontal and vertical
- If you select a tribble or df, reflows it pretty-like


# When upgrading R

If you upgrade R to a new major or minor release (eg, 1.3 or 2.0), you will lose all packages that are installed!

ALWAYS install new versions of R using the installer from the official R page—DO NOT USE HOMEBREW

BEFORE upgrading, get a list of all your packages:

```{r eval = FALSE}
devtools::package_info("installed")
```

This will give a df showing package names, versions, and where they're from (CRAN or github).  Save this somewhere so you can read it back in after you install the upgrade.

Next, get a list of which addins you have active from rsam.  Run rsam::rsam(), open it in Safari, and then copy/paste the list into TextEdit.  You'll have to manually re-enter all this stuff using rsam::rsam() after the upgrade.

Now, upgrade.

Then after installing, filter by source and install as appropriate.  Will need to install `devtools` first

```{r}
CRAN_packages <- devtools::package_info("installed") %>% filter(str_detect(source, "CRAN"))
github_packages <- devtools::package_info("installed") %>% filter(str_detect(source, "Github"))
```

```{r eval = FALSE}
install.packages(CRAN_packages$package)
install_github(github_packages$package)
```




# Stuff to add

(broom)[https://broom.tidymodels.org] to tidy up models


```{r}
a <- b <- matrix(1:100, ncol=2)
a <- a[-20,]
b <- b[-45,]
b[c(18, 44)] <- 999
diffobj::diffPrint(target=a, current=b)
```

Nesting - unnest, unnest_longer, unnest_wider, hoist (tidyr)

https://uribo.github.io/rpkg_showcase/index.html


http://personality-project.org/r/psych/

https://github.com/eddelbuettel/anytime VS lubridate

zoo - rollmean(), rollapply()...

lobstr package

iris %>% 
  boomer::boom()

https://github.com/Pakillo/grateful

https://cran.r-project.org/web/packages/report/vignettes/report.html

https://cran.r-project.org/web/packages/MBESS/index.html


https://rpkgs.datanovia.com/ggpubr/index.html