---
title: "Useful R Stuff"
output:
  html_document:
    css: "style.css"
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false

---
<script src="fold_rmd.js"></script>


<!-- <div class = "fold s o"></div> to hide "s"ource or "o"utput or both -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r echo = FALSE, results = "hide", message = FALSE}
library(tidyverse)

```


# Working with dataframes
## janitor - clean dirty data, tabulating tools, adornments

Clean up column names
```{r results = "hide", message = FALSE}
library(janitor)
```

```{r}
# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")

test_df %>%
  clean_names()

test_df %>% 
  clean_names(case = "small_camel")
```

Compare two or more DFs to see if they have the same types of columns and can be rbound
```{r}
df1 <- data.frame(a = 1:2, b = c("big", "small"))
df2 <- data.frame(a = 10:12, b = c("medium", "small", "big"), c = 0, stringsAsFactors = TRUE) # here, column b is a factor
df3 <- df1 %>%
  dplyr::mutate(b = as.character(b))

compare_df_cols(df1, df2, df3)
```

### tabyls - compare combinations of varibales like xtab

```{r}
mtcars %>%
  tabyl(gear, cyl)
```

You can "adorn" tabuls with totals, percentages, and more

```{r}
mtcars %>%
  tabyl(gear, cyl) %>%
  adorn_totals("col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns() %>%
  adorn_title()
```

[More on tabyls](http://sfirke.github.io/janitor/articles/tabyls.html)



## diffobj - compare two dfs (or other objects) to see what's changed between them

NB: in RStudio this will open a colorful thing in the Viewer that's easy to read

<div class = "fold o">
```{r}
library(diffobj)
a <- b <- matrix(1:100, ncol=2)
a <- a[-20,]
b <- b[-45,]
b[c(18, 44)] <- 999
diffPrint(target=a, current=b)
```
</div>


<div class = "fold o">
```{r}
diffChr(letters[1:3], c("a", "B", "c"))
```
</div>



# Summarizing
## skimr::skim() - nice summary statistics for dfs and the like

<div class = "fold o">
```{r}
library(skimr)
iris %>% 
  group_by(Species) %>% 
  skim() %>% 
  print() # print() only necessary so we don't get a kable table in this document
```
</div>

---

```{r}
iris %>% 
  skim() %>% 
  summary() %>% 
  print() # print() only necessary so we don't get a kable table in this document
```

---

<div class = "fold o">
```{r}
iris %>% 
  skim(Sepal.Length, Petal.Length) %>%  # tidyselect columns to summarize
  print() # print() only necessary so we don't get a kable table in this document
```
</div>

---

<div class = "fold o">
You can do cool stuff in the pipeline with it

```{r}
iris %>%
  skim() %>%
  dplyr::filter(numeric.sd > 1) %>% # "numeric" is the variable type that we're targeting
  print()
```

</div>

# Plotting
## Khroma - color scales that are nice and colorblind friendly

https://packages.tesselle.org/khroma/


```{r}
library(khroma)
bright <- colour("bright")
bright(7)
```

<div class="fold o">
```{r}
# Preview the theme
plot_scheme(bright(7), colours = TRUE)
```
</div>

Use with ggplot2

```{r}

data(mpg, package = "ggplot2")

ggplot2::ggplot(data = mpg) +
  ggplot2::aes(x = displ, y = hwy, colour = class) +
  ggplot2::geom_point() +
  scale_colour_bright()
```



## ggiraph - interactive graphs even in Rmd

This can do a whole lot of cool stuff more than this simple example.

More info: <https://davidgohel.github.io/ggiraph/>


```{r}
library(ggplot2)
library(ggiraph)
data <- mtcars
data$carname <- row.names(data)

gg_point = ggplot(data = data) +
    geom_point_interactive(aes(x = wt, y = qsec, color = disp,
    tooltip = carname, data_id = carname)) + 
  theme_minimal()

girafe(ggobj = gg_point)
```



## Patchwork - composing multiple plots together

Setting up some plots we can stitch together
<div class = "fold s">
```{r messages = FALSE}
library(patchwork)
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```
</div>


<div class = "fold o">

```{r}
p1 + p2 # combine two plots
```
</div>


<div class = "fold o">
```{r}
p1 + p2 + p3 + p4 # by default it makes a square grid, filling rowwise
# plot_layout()
p1 + p2 + p3 + p4 + plot_layout(nrow = 3, byrow = FALSE)
```
</div>


Stacking and packing plots:
<div class = "fold o">
```{r}
p1 / p2 # / stacks one on top of the other
p1 | (p2 / p3) # | plots side by side.  Note the parens
```
</div>

<div class = "fold o">
```{r}
p1 + p2 +
  plot_annotation(
    title = 'This is a title',
    caption = 'made with patchwork',
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5))
  )
```
</div>

You can define your own layout using letters, with relative sizing
<div class = "fold o">
```{r}
layout <- "
##BBBB
AACCDD
##CCDD
"
p1 + p2 + p3 + p4 + 
  plot_layout(design = layout)
```

</div>




It gets a lot more [complicated and cool](https://patchwork.data-imaginist.com/).  See [here](https://patchwork.data-imaginist.com/articles/guides/multipage.html) for how this package can also avoid plots "jumping" around as you switch slides.

# Package and project management


## Package development workflow
<https://r-pkgs.org>: the definitive book by Wickham and Bryan, which explains lots of the stuff below

```{r eval = FALSE}

devtools::load_all() # simulates installing and loading your package
devtools::document() # updates all the documentation and stuff via roxygen2
devtools::check() # run some checks when you think you're ready to release

```


## usethis


```{r eval = FALSE}
# Creating
create_package()
create_project()

# Checking
proj_sitrep() # show wd, proj dir, etc

# Development
use_package() # import packages into DESCRIPTION

# Include important things
use_pipe()
use_tibble()

# Git
use_git()
use_github()
git_sitrep()

```
Also has git/github functions and other config stuff



# Modeling

## TidyModels
<https://www.tidymodels.org>; has a bunch of subpackages, including broom



# Data exploration
## DataExplorer
<https://boxuancui.github.io/DataExplorer/>

<div class = "fold o">
```{r message=FALSE}
library(DataExplorer)
plot_bar(ggplot2::diamonds) # Plot frequency of levels in each factor
```
```{r}
plot_histogram(ggplot2::diamonds) # Histogram for all continuous variables
```
```{r}
plot_boxplot(ggplot2::diamonds, by = "cut")
```
</div>

# Stuff to add

(broom)[https://broom.tidymodels.org] to tidy up models


```{r}
a <- b <- matrix(1:100, ncol=2)
a <- a[-20,]
b <- b[-45,]
b[c(18, 44)] <- 999
diffobj::diffPrint(target=a, current=b)
```





