---
title: "Upgrading R without losing all your packages"

date: "6-30-23"

categories:
  - upgrading
  - package management
---

{{< include /_include_header.qmd >}}

<!-- begin content -->

If you upgrade R to a new major or minor release (e.g. 4.0 or 4.1, but not 4.1.1), your library of installed packages doesn't come along for the ride.

## Before you upgrade

*Before* you upgrade, you should make a list of all of your installed packages---this is easy with `sessioninfo::package_info("installed")`. If you've got packages installed from multiple sources (like CRAN and also Github), it's best to save these as separate lists to facilitate automated reinstallation of these grouped by source:

```{r}
#| messages: false

library(tidyverse)

# Figure out what's installed
installed_packages <- sessioninfo::package_info("installed") |> 
  mutate("where_from" = dplyr::case_when(
    str_detect(source, "CRAN") ~ "CRAN",
    str_detect(source, "Github") ~ "Github",
    .default = "other" # catch any packages from other sources;
                       # not automatically installed below
    )
  ) |> 
  nest(.by = where_from)

# Where to write the lists of packages.  Change as needed
file_location <- fs::path("~/Desktop/tmp/")

# Write these lists to disk
installed_packages %>% 
  pwalk(function(where_from, data) {
    write_csv(data,
              file = fs::path(file_location,
                              paste0(where_from, "_packages"),
                              ext = "csv"))
  })

```


## After you upgrade 

Then, after you upgrade R, you can re-install the packages from these lists:

```{r}
#| eval: false

# Install manually some things we'll need since they won't be installed anymore
install.packages(c("remotes", "dplyr", "stringr"))

file_location <- "~/Desktop/tmp/" # Change as needed

# Install CRAN packages
CRAN_packages <- read.csv(paste0(file_location, "CRAN_packages", ".csv"))
remotes::install_cran(CRAN_packages$package)

# Install Github packages
github_packages <- read.csv(paste0(file_location, "Github_packages", ".csv")) |>
  dplyr::mutate(github_source = stringr::str_extract(source,
      "(?<=Github \\().*(?=@)")) # Get the username and repo name
remotes::install_github(github_packages$github_source)
```
NB: If you have locally installed packages, or packages from another source, these aren't automatically reinstalled with the code above, but are listed in `other_packages.csv`.

<!-- End content -->

{{< include /_include_footer.qmd >}}
