---
title: "`khroma` and `colorblindr`: Tools for colorblind-friendly plots"

date: "7/4/23"

categories:
  - visualization
  - colors

toc: true

---

{{< include /_include_header.qmd >}}

<!-- begin content -->

```{r setup}
#| code-fold: true
#| code-summary: "Setup/libraries"
#| output: false
#| warning: false

library(dplyr)
library(ggplot2)
```

```{r}
#| include: false

theme_set(asbutils::theme_asb())
```


## `khroma` - colorblind-friendly palettes

[`khroma`](https://packages.tesselle.org/khroma/) contains a number of color schemes that are colorblind-friendly, divided into diverging, qualitative, and sequential types:

```{r}
khroma::info() |>
  group_by(type) |>
  summarize(palettes = paste(palette, collapse = ", "))
```

The `color` function takes a palette name and returns a function, which in turn returns a palette of specified size:

```{r}
library(khroma)

bright_fun <- color("bright")
bright_fun(5)
```
And you can preview these palettes with `plot_scheme()`:

```{r}
#| fig-height: 2

plot_scheme(bright_fun(7), names = TRUE,
            colours = TRUE) # NB: `colours` must be spelled British-ly
```

Further, you can preview the palette as viewed by several types of colorblindness:

```{r}
#| classes: .fold .o

plot_scheme_colorblind(bright_fun(7))
```
<br>

Each of the schemes have built-in `ggplot` scales in the form `scale_color_{palettename}` and `scale_fill_{palettename}`:

```{r}
#| fig-cap: Percent of population with a college degree vs. percent of population below the poverty line, by state.  Data from `ggplot2::midwest`
#| fig-alt: Percent of population with a college degree vs. percent of population below the poverty line, by state.  Data from `ggplot2::midwest`

plot <- midwest |> 
  filter(percbelowpoverty < 30) |> 
  ggplot(aes(x = percollege, y = percbelowpoverty, color = state)) +
    geom_point(size = 3, alpha =.5) +
    scale_color_bright()

plot
```


## `colorblindr` - Preview your plots with colorblind simulation

`colorblindr::cvd_grid` will simulate how your plots will look to people with various types of colorblindness.  The `midwest` plot from above is shown here:

```{r}
#| warning: false
#| fig-width: 10

# remotes::install_github("clauswilke/colorblindr")

colorblindr::cvd_grid(plot)
```


<!-- End content -->
{{< include /_include_footer.qmd >}}

