---
title: "Paginate a faceted graph and save as a single multi-page PDF"

date: "7/1/23"

categories:
  - visualization
---

{{< include /_include_header.qmd >}}

<!-- begin content -->

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(ggforce)
ggplot2::theme_set(asbutils::theme_asb())
```

When you have a plot with just a few facets, it's easy to fit them all on a single page. But what if you have a whole bunch of facets that you want to visualize? For example, the `USArrests` dataset has data from all 50 US states.

If we just use the usual `facet_wrap` or `facet_grid` from `ggplot2`, you end up with one massive, smooshed-together image:

```{r}
#| message: false

library(tidyverse)

plot_1 <- USArrests |>  # dataset on US arrests
  rownames_to_column("State") |>   # This is just so we can see the state names
  ggplot(aes(x = UrbanPop, y = Murder)) +
  geom_point()  +
  facet_wrap(~ State,
             nrow = 5, ncol = 10)

plot_1
```

The [`ggforce`](https://ggforce.data-imaginist.com) library, among its many useful functions, includes `facet_wrap_paginate` and `facet_grid_paginate`. These work like `facet_wrap` and `facet_grid`, but take an additional `page` argument. Here we specify that we want 4 rows and 5 columns per page, so for 50 states there will be 2 full pages plus one page with the remaining 10 states:

```{r}
library(ggforce)

plot_2 <- plot_1  +
  facet_wrap_paginate(~ State,
             nrow = 4, ncol = 5, page = 2)

plot_2
```

Note that we had to specify which page we wanted to print. We can then easily put this into a loop to generate all the necessary pages. The helpful `n_pages()` will count the pages needed, which we can then use in a loop. This gives us 3 separate plots.

```{r}
#| echo: false
#| include: false

ggplot2::theme_set(asbutils::theme_asb())
```


```{r}
#| classes: .fold .o .g

plot_pages <- plot_2 |> 
  n_pages()

for (i in 1:plot_pages) {
  print(
    plot_1 + facet_wrap_paginate(~ State,
             nrow = 4, ncol = 5, page = i)
  )  
}

```

<br>

That's better, but for inclusion in a longer document or for sharing the plot, you may want all three pages in a single PDF. This can be accomplished by calling `pdf()` before the loop and `dev.off()` after.^[In the past I've had issues with getting `pdf()` to work, especially if I use any fancy fonts in the ggplot theme.  If you run into issues, you can replace `pdf()` with `cairo_pdf(..., onefile = TRUE)`, which seems to work even with exotic fonts.]

```{r}
#| eval: false
#| output: false

pdf('many_plots.pdf', width = 11, height = 8.5) #start building pdf
  for (i in 1:plot_pages) {
    print( # don't forget this
      plot_1 +
      facet_wrap_paginate(~State, nrow = 4, ncol = 5, page = i)
    )
  }
dev.off() # end building pdf
```

The [resulting PDF](many_plots.pdf) is a single file with 3 pages, as we expect.

(This post was inspired by [this conversation on StackOverflow](https://stackoverflow.com/questions/48544039/how-to-save-output-from-ggforcefacet-grid-paginate-in-only-one-pdf).)

<!-- End content -->

{{< include /_include_footer.qmd >}}
