---
title: "`janitor`: Clean dirty data, plus improved tables and crosstab"

date: 6/25/23

categories:
  - data cleaning

toc: true
---

{{< include /_include_header.qmd >}}

<!-- begin content -->

```{r setup}
#| code-fold: true
#| code-summary: "Setup"
#| output: false
#| warning: false

library(janitor)
```


[`janitor`](https://sfirke.github.io/janitor/index.html) contains various tools for examining and cleaning dirty data.


## Cleaning dirty data

### Clean column names

Let's create a `df` with some poorly-chosen column names:

```{r}
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")

```

`clean_names()` does just as the name implies:

```{r}
test_df |> 
  clean_names()
```


The `case` argument to `clean_names()` specifies what case you'd like output names to be in.  You can specify any case style that's available in `snakecase::to_any_case()`, including "screaming_snake" if you want to be perverse:

```{r}
test_df |> 
  clean_names(case = "screaming_snake")
```

### Check if `df`s are row-bind-able

Also useful is `compare_df_cols` which summarizes whether the specified `df`s can be row-bound (i.e., have columns of the same names/types):

```{r}
df1 <- data.frame(A= 1:2, b = c("big", "small"))
df2 <- data.frame(a = 10:12, b = c("medium", "small", "big"), c = 0, stringsAsFactors = TRUE) # here, column b is a factor
df3 <- df1 |> 
  dplyr::mutate(b = as.character(b))

compare_df_cols(df1, df2, df3)
```

If you just want a simple `TRUE`/`FALSE` value telling you whether the `df`s match, you can use `compare_df_cols_same()`:

```{r}
str(compare_df_cols_same(df1, df2, df3, verbose = FALSE))
```



## Examining data and crosstabs

`janitor`'s version of tables are called `tabyls`.  You can easily generate crosstabs:

```{r}
palmerpenguins::penguins |> 
  tabyl(species, island)
```

There are lots of ways to pretty up the output via `adorn_*` commands, giving things like column or row percentages, optionally with `n`s in parentheses:

```{r}
palmerpenguins::penguins |> 
  tabyl(species, island) |> 
  adorn_totals("col")  |> 
  adorn_percentages("row")  |> 
  adorn_pct_formatting(digits = 2) |> 
  adorn_ns()
```
You can also use the `adorn_*` functions on regular ol' `df`s:

```{r}
palmerpenguins::penguins |> 
  dplyr::sample_n(10) |> # <1>
  dplyr::select(-year) |> 
  adorn_totals("row")
```
1. Choose 10 random rows so this doesn't print forever

Check out the [`tabyl`s vignette](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html) for more info.

<!-- End content -->
{{< include /_include_footer.qmd >}}

