---
title: "`thinkr::peep()`: See intermediate stages of piped dfs"

date: "6/1/23"

categories:
  - dataframes
  - piping
---

{{< include /_include_header.qmd >}}

<!-- begin content -->

```{r setup}
#| code-fold: true
#| code-summary: "Setup"
#| output: false
#| warning: false

library(tidyverse)
library(thinkr)
```

Insert `thinkr::peep(head)` into a `df` pipeline to print a few rows at a particular intermediate stage in the pipeline. Here it prints a few lines of the `df` before it gets summarized:

```{r}
palmerpenguins::penguins |> 
  peep(head) |>
  group_by(species) |>  # <1>
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

1.  How convenient that the base R pipe looks like a penguin.

You can use multiple functions, like `head` and `tail` to see two views of the intermediate step; `verbose = TRUE` prints the functions used for these peeps:

```{r}
palmerpenguins::penguins |>
  peep(head, tail, verbose = TRUE) |> 
  group_by(species) |> 
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

And you can include arguments to functions inside `peep()` with dot syntax if you're using the `magrittr` pipe `%>%` (at least for the line directly above the `peep`):^[You can't use the base pipe's placeholder, `_`, here, since it does not work in nested functions.]

```{r}
palmerpenguins::penguins %>% 
  peep(head(n = 7, x = .)) |> 
  group_by(species) |> 
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

<!-- End content -->
{{< include /_include_footer.qmd >}}